---
title: "Plot MeBr Apple Fumigation Data"
author: "John Maindonald"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r linkfun, message=FALSE}
library(qra, quietly=TRUE)
require(ggplot2)
```

## Plot data
```{r plot}
codling <- DAAG::codling
num1 <- match(unique(codling$gp),codling$gp)
rnum <- rep(1, nrow(codling))
rnum[num1] <- 2
nnew <- rep(1:nrow(codling),rnum)
nctl <- match(num1,nnew)
codling0 <- codling[nnew,]
codling0[nctl,c('dose','tot','dead','pobs','cm','ct')] <-
  with(codling0[nctl,], cbind(rep(0,length(num1)), numcm, round(cm*numcm), 
       cm,rep(NA,length(num1)),rep(0,length(num1))))     

codling0[["logCT"]] <- log(codling0[["ct"]])
codling0[["Rep"]] <- with(codling0, qra::gpsWithin(gp, list(Cultivar,year)))
```

# Plot 1988 and 1989 data separately
```{r y1988, fig.width=7, fig.height=2.5, eval=TRUE, out.width='100%'}
library(ggplot2)
graphSum(df=codling0, subSet=expression(year==1988),
                     link="cloglog", logScale=TRUE,
                     dead="dead", tot="tot", dosevar="logCT", Rep="Rep",
                     fitRep=NULL, fitPanel=NULL,
                     byFacet=~Cultivar,
                     maint="Codling moth, MeBr",
                     xlab=expression(bold("CT ")*"(gm.h."*m^{-3}*")"))
```

```{r y1989, fig.width=6.5, fig.height=2.5, out.width="75%", eval=TRUE}
graphSum(df=codling0, subSet=expression(year==1989),
                     link="cloglog", logScale=TRUE,
                     dead="dead", tot="tot", dosevar="logCT", Rep="Rep",
                     fitRep=NULL, fitPanel=NULL,
                     byFacet=~Cultivar,
                     maint="Codling moth, MeBr",
                     xlab=expression(bold("CT ")*"(gm.h."*m^{-3}*")"))
```

## GLM: quasibinomial link - no ctl mortality adjustment

```{r GLMquasi}
cloglog <- stats::make.link("cloglog")[["linkfun"]]
redDel <- subset(codling,
                 Cultivar=="Red Delicious")
redDel <- droplevels(redDel)
   redDel.glm <- glm(pobs~ct, data=redDel, weights=tot,
                     family=quasibinomial(link='cloglog'))
   vv <- summary(redDel.glm)$cov.scaled
   ci <- fieller(0.99, b=coef(redDel.glm), vv=vv, link='cloglog')
print(round(ci,2))
```   

```{r yr88, out.width="30%", warning=FALSE}
library(splines, quietly=TRUE)
codling88  <- droplevels(subset(DAAG::codling, year==1988))
cult <- sort(unique(codling88$Cultivar))
for(culti in cult){
dset <- subset(codling88, Cultivar==culti)
gp <- unique(dset$gp)
k <- 1
for (j in gp){
  k <- k+1
  dsetj <- droplevels(subset(dset, gp==j))
  gcm <- cloglog(dsetj[['cm']][1])
  ylim=range(c(gcm,cloglog(dsetj[['pobs']])))
  if(j==gp[1]) plot(cloglog(pobs)~log(ct), data=dsetj, col=k, ylim=ylim) else 
    points(cloglog(pobs)~log(ct), data=dsetj, col=k)
  abline(h=gcm, col=k, lty=3)
dset1.glm <- glm(pobs~log(ct), data=dsetj, weights=tot,
                   family=quasibinomial(link='cloglog'))
dset2.glm <- glm(pobs~ns(log(ct),2), data=dsetj, weights=tot,
                   family=quasibinomial(link='cloglog'))
fit1 <- predict(dset1.glm)                   
fit2 <- predict(dset2.glm)
lines(log(dsetj[['ct']]),fit1, col=k)
lines(log(dsetj[['ct']]),fit2, col=k, lty=2)
}
mtext(side=3, line=0.5, culti)
}
```

```{r yr89, fig.asp=1, out.width="30%"}
library(splines, quietly=TRUE)
codling89  <- droplevels(subset(DAAG::codling, year==1989))
cult <- sort(unique(codling89$Cultivar))
for(culti in cult){
dset <- subset(codling89, Cultivar==culti)
gp <- unique(dset$gp)
k <- 1
for (j in gp){
  k <- k+1
  dsetj <- droplevels(subset(dset, gp==j))
  gcm <- cloglog(dsetj[['cm']][1])
  ylim=range(c(gcm,cloglog(dsetj[['pobs']])))
  if(j==gp[1]) plot(cloglog(pobs)~log(ct), data=dsetj, col=k, ylim=ylim) else 
    points(cloglog(pobs)~log(ct), data=dsetj, col=k)
  abline(h=gcm, col=k, lty=3)
dset1.glm <- glm(pobs~log(ct), data=dsetj, weights=tot,
                   family=quasibinomial(link='cloglog'))
dset2.glm <- glm(pobs~ns(log(ct),2), data=dsetj, weights=tot,
                   family=quasibinomial(link='cloglog'))
fit1 <- predict(dset1.glm)                   
fit2 <- predict(dset2.glm)
lines(log(dsetj[['ct']]),fit1, col=k)
lines(log(dsetj[['ct']]),fit2, col=k, lty=2)
}
mtext(side=3, line=0.5, culti)
}
```

```{r byRep2, out.width="30%"}
codling <- DAAG::codling
codling$cultYear <- with(codling, paste0(Cultivar, year))
codling$Rep <- with(codling, qra::gpsWithin(gp,cultYear))
add2.glm <- glm(pobs~cultYear*Rep+ns(log(ct),2), data=codling,
                family=quasibinomial(link='cloglog'))
byRep.glm <- glm(pobs~cultYear*Rep*ns(log(ct),2), data=codling,
                 family=quasibinomial(link='cloglog'))
anova(add2.glm, byRep.glm, test='F')
lab <- paste(1:nrow(codling))
plot(add2.glm, which=1:3, labels.id=lab)
```

## Repeat, omitting observations 72 and 91

In the analysis that omits rows 72 and 91, row 70 then appears as
an outlier. All these are points that lie above the fitted curve.

```{r omit72and91}
add2.glm <- glm(pobs~cultYear*Rep+ns(log(ct),2), data=codling[-c(72,91),],
                family=quasibinomial(link='cloglog'))
byRep.glm <- glm(pobs~cultYear*Rep*ns(log(ct),2), data=codling[-c(72,91),],
                 family=quasibinomial(link='cloglog'))
anova(add2.glm, byRep.glm, test='F')
add2.glm <- glm(pobs~cultYear*Rep+ns(log(ct),2), data=codling[-c(70,72,91),],
                family=quasibinomial(link='cloglog'))
byRep.glm <- glm(pobs~cultYear*Rep*ns(log(ct),2), data=codling[-c(70,72,91),],
                 family=quasibinomial(link='cloglog'))
anova(add2.glm, byRep.glm, test='F')
```

It is interesting that, in contrast to data from more recent work, 
the scale-location plot is flat, indicating that dispersion is close
to constant over the whole of the range.

## Random effects analysis

```{r glmer}
library(lme4)
codling$live <- with(codling, tot-dead)
codling$sclogdose <- with(codling, scale(log(dose)))
add2.glmer <- glmer(cbind(dead,live)~cultYear*ns(sclogdose,2)+
                    (sclogdose|gp), data=codling, nAGQ=0,
                    family=binomial(link='cloglog'))
```
