---
title: "Quantal Response Analysis Functions"
author: "John Maindonald, Statistics Research Associates"
date: "`r format(Sys.Date(),'%d %B %Y')`"
documentclass: article
classoption: b5paper
fontsize: 10pt
output:
  bookdown::html_document2:
    theme: cayman
    highlight: vignette
    base_format: prettydoc::html_pretty
    toc: true
    toc_depth: 2
    number-sections: true
    pandoc_args: NULL
    link-citations: true
bibliography: qra.bib
vignette: >
  %\VignetteIndexEntry{Quantal Response Analysis Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \usepackage{xcolor}
---

<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0;
}
body{
   font-size: 14pt;
}
td{
  font-size: 14pt;
}
code.r{
  font-size: 13pt;
}
pre {
  font-size: 13pt
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment=NA, width=70)
options(show.signif.stars=FALSE)
```

# Introduction {-}


Quantal responses are counts of all-or-none responses,
out of a total $n$.  A dose-response relationship 
quantifies the response as a function of dose, 
or more generally as a function of an exposure.  

Data are from a broad class of experiments where responses 
are insect deaths out of some total number exposed. Exposure
may be time in coolstorage, or dose of a fumigant, or a
concentration-time measure of exposure to a fumigant, or an
intensity-time measure of exposure to radiation. 
See @follett2006current for commentary on the regulatory 
and scientific background.  We will use Hawaian fruitfly data that 
has been supplied by Dr Peter Follett to demonstrate the use of 
functions in the _qra_ package, where the exposure is time in
coolstorage,

The following code sets up the data.

```{r prepareData, echo=TRUE}
HawCon <- qra::HawCon
## Change name "CommonName" to "CN", for more compact output.
CCnum <- match("CommonName", names(HawCon))
names(HawCon)[CCnum] <- "CN"
## trtGp will identify species & lifestage combination
## trtGpRep will identify species, lifestage, and rep
## cTime is centered version of TrtTime
## scTime is centered and scaled version of TrtTime,
## needed to get some mixed model fits to converge
HawCon <- within(HawCon, {
  trtGp <- factor(paste0(CN,LifestageTrt, sep=":"))
  trtGpRep <- paste0(CN,LifestageTrt,":",RepNumber)
  scTime <- scale(TrtTime)
  obs <- factor(1:nrow(HawCon))
})
```

# Data setup and choice of model

For the data that will be considered here, the exposure 
measure is time in coolstorage, and the response is 
mortality of insect pests. 

## Graphical Display

```{r cap1, echo=FALSE}
cap1 <- " Graphs are designed to give an indication of the pattern, 
when mortalities are shown on a complementary log-log scale, of 
mortality response with days in coolstorage."
```

```{r plots, fig.width=7, fig.height=7.5, fig.align='center', out.width="75%", fig.cap=cap1}
qra::graphSum(df=HawCon, link="cloglog", logScale=FALSE,
         dead="Dead", tot="Total", dosevar="TrtTime",
         Rep="RepNumber", fitRep=NULL, fitPanel=NULL,
         byFacet=~trtGp, layout=LifestageTrt~Species,
         xlab="Days", maint="Hawaian contemporary data")
```

Figure \@ref(fig:plots) is designed to give a broad indication
of the pattern of response, on a complementary log-log link scale.
Responses appear acceptably linear, at least after the first one 
or two observations.  There are clear systematic differences 
between replicates, indicative of a strong between replicate 
component of variation.

## The choice of link function

Commonly used link functions are 

* `logit`: This is difficult to distinguish, 
for any except datasets that have very large numbers of 
observations at exposures that lead to very low or very 
high mortality, from the `probit`.  With the logit link,
the model is predicting the $\log(\mbox{odds})$ of response.
The `probit` link can be motivated by assuming the presence
of a normally distributed random variable that generates
a response whenever its value crosses a 
threshold.^[<See https://en.wikipedia.org/wiki/Probit_model>,
accessed 29 May 2021]
* `complementary log-log`, abbreviated to `cloglog`: 
This arises naturally as an extreme value distribution, when
(for insect mortality), death arises from the failure of
any one of a number of organs.

While later analyses will suggest a mild preference for
a complementary log-log link function, as against a
logit link, the difference is small.  The difference
does matter, for extrapolation to  mortality values
(commonly 99% or greater) that lie close to or beyond 
the limits of the data. The fitted models should, 
for this purpose, be treated as providing an indication 
of the broad pattern of response.

The possible use of a transformation for the exposure variable 
(or variables), commonly a logarithmic transformation, gives 
further flexibility.  One or other choice within the range of
possibilities noted has often been found to work well in practice,
at least to the extent that the model survives
scrutiny under standard diagnostic checks. 
Where the model will be used to make predictions beyond the 
limits of the main body of the data, this adds uncertainty.

## Modeling the error distribution

With data such as here, it is to be expected that the response 
will show strong extra-binomial variation.  This happens because 
the response varies from insect to insect, and/or because 
insects do not respond independently.  The `glmmTMB` package 
[@glmmTMB] implements the `betabinomial` error family, with
the option to model the scale parameter, and hence the 
multiplier for the binomial variance, as a function of
explanatory variable(s).  In the terminology used for R's 
`glm()` function, and that will be used in the sequel, the
multiplier for the binomial variance is referred to as the 
dispersion factor.  For the data considered here, this 
dispersion factor is high in the middle of the range of data, 
reducing at the extremes.

An alternative that will be investigated in a later section
is the use of binomial errors, with observation level random effects
used to account for differences at the level of individual
observations.  For the present data, this automatically
achieves much the same effect as betabinomial errors,
without the need for specific attention to the modeling
of a dispersion factor.  The model fit appears, however
to be less satisfactory than the use of betabinomial errors
with a dispersion factor adjustment.

Other error models that are described in the literature, 
and that operate at the level of individual observations,
are discussed in the vignette 
`vignette("countDists", package = "qra")`.

## Choices required for mixed model fits

Fits will require a choice of link functions, modeling of
the fixed effects, and modeling of the error distribution.
Because there are just three replicates per lifestage,
it is necessary to base estimates on a model that brings
together components of variance information across
lifestages.  This inevitably leads to estimates that will
sometimes smooth out effects that are specific to an 
individual lifestage.

The subsection that follows will explore the use of a 
betabinomial error model, as implemented in the 
__glmmTMB__ package. The parameterization is that 
described in @morris1997disentangling, with parameters 
$\mu$ and $\phi$.  Here $\mbox{E}[x] = n \mu$, and
$$
\mbox{var}[x] = n \mu (1-\mu) \dfrac{\phi+n}{\phi+1}
$$
Setting $\phi = \frac{1-\rho}{\rho}$, where $\rho$ is 
the intra-class correlation, this equals
$$
n \mu(1-\mu)(1+(n-1)\rho)
$$

### Quasibinomial errors {-}

The `lme4::glmer()` function offers the option of a quasibinomial 
error, as for `stats::glm()`.  It does not offer an equivalent to 
the `glmmTMB` dispformula.

Specification of a quasibinomial error has the consequence that
the model is fitted as for a binomial distribution, with the
the binomial variance $n \pi (1- \pi)$ then multiplied by a
constant factor $\Phi$ that is usually estimated using the
Pearson chi-squared statistic.  Compare this with the betabinomial, 
where the multiplier is $\Phi = 1+(n-1)\rho$, i.e., it increases 
with $n$.  This is an important difference.

## Complementary log-log versus logit link

Figure \@ref(fig:glmmTMB-altFits-gph1)  shows lines and curves
from alternative models that have been fitted to the data.

```{r echo=FALSE}
pkg <- "glmmTMB"
pcheck <- suppressWarnings(requireNamespace(pkg, quietly = TRUE))
if(pcheck) pcheck & packageVersion("glmmTMB") >= "1.1.2"
if(!pcheck){
  message("`glmmTMB` version >= 1.1.2 is not available")
  message("Will not continue execution of vignette")
  knit_exit()
  }
```

```{r glmmTMB-altFits, message=FALSE, warning=FALSE, echo=FALSE}
if("VGAM" %in% (.packages()))
  detach("package:VGAM", unload=TRUE)
# Load packages that will be used
suppressMessages({library(lme4); library(splines)})
form <- cbind(Dead,Live)~0+trtGp/TrtTime+(1|trtGpRep)
form2s <- cbind(Dead,Live)~0+trtGp/TrtTime+ns(scTime,2)+(1|trtGpRep)
HCbb.cll <- glmmTMB::glmmTMB(form, dispformula=~trtGp+ns(scTime,2), 
                    family=glmmTMB::betabinomial(link="cloglog"),
                    data=HawCon)
HCbb2s.cll <- update(HCbb.cll, formula=form2s)
HCbb.logit <- glmmTMB::glmmTMB(form, dispformula=~trtGp+ns(TrtTime,2), 
                      family=glmmTMB::betabinomial(link="logit"),
                      data=HawCon)
HCbb2s.logit <- update(HCbb.logit, formula=form2s)
```

One has to experiment to get these models to fit.  Some of the 
possibilities are:

* Stay with default control parameters, or try an alternative, 
such as `ctl <- glmmTMBControl(optimizer=optim, optArgs=list(method="BFGS"))`
* Replace `TrtTime` by `scTime`, which is the centered and
scaled version, in some or all of
    + the random effects term(s) in the model formula  
    + the fixed effects terms  
    + the dispersion formula (`dispformula`).
* As will be shown later, the dispersion formula for the 
`HawCon` data needs to be able to fit a curve that has a roughly
hill shape.  Degree 2 normal splines (`splines::ns(x,2)` , where
in our case `x=scTime`) appear to work reasonably well.
* Use of `update()` to update a simpler model (e.g., to add the 
degree 2 term in models that added a curvature term to the
straight line in the fixed effect) may avoid warning messages 
or failures that otherwise result. 

Normal spline curves, or polynomial curves, may be used to 
model the fixed effects, as alternatives to lines. This can 
be compared with fitting a line, in order to check for 
curvature in the response.

If `glmer()` (from the `glmer` package) is used in place of
`glmmTMB()`, the `lme4` function `allFit()` can be used to
compare results between a range of available optimizers.

Both the complementary log-log model and the logit model appear
to fit well, as judged by examining plots of randomized quantile
residuals.

### Details of the model fitting process {-}

Code that fits the several models is:
```{r glmmTMB-altFits, eval=FALSE, echo=-(1:2)}
```

The right hand side of the model formula divides into two parts:

* The first part, i.e., `0+trtGp/TrtTime`, expands to
`0 + trtGp + trtGp:TrtTime`.  This specifies a different
constant term and different slope, and thus a different 
line, for each different treatment group.
    + If the `0` is omitted, so that this initial part of
    the formula reduces to `trtGp/TrtTime`, all that
    changes is the parameterization.  There is then an
    overall constant term, with treatment group effects
    expressed as differences from the overall constant term.
* The round brackets that enclose the remainder of the 
  right hand side of the formula, i.e., `(scTime|trtGpRep)`, 
  identify it as supplying  random effects terms.  Here, a
  different random effects line is added for each replicate
  (`trtGpRep`).  This is achieved by fitting a random
  intercept and a random slope for each different replicate.
    + Note that `scTime` is by default interpreted as 
    `1 + scTime`.

In addition, all models use a `dispformula` to allow for 
change in the betabinomial scale parameter $\phi$. The
`dispformula` used, allowing a different degree 2 function 
of `scTime` for each different treatment group, was chosen
after some experimentation. 

## Fitted lines, vs fitted normal spline curves

Figure \@ref(fig:glmmTMB-altFits-gph1) shows fitted lines, 
and fitted degree 2 normal spline curves, in Panel A for a
complementary log-log (cloglog) link, and in Panel B for 
a logit link.

```{r cap7, echo=FALSE}
cap7 <- "Differences are shown, between fitted degree 2 normal spline 
curves and fittes lines.
Panel A is for the models that use a complementary log-log
(cloglog) link, while Panel B is for a logit link."
```

```{r glmmTMB-altFits-gph1, fig.width=9, fig.height=4.0, fig.show='hold', top=2, out.width="100%",  fig.align='center', message=FALSE, warning=0, fig.cap=cap7, echo=FALSE}
library(lattice)
clog <- make.link('cloglog')$linkfun
logit <- make.link('logit')$linkfun
cloginv <- make.link('cloglog')$linkinv
logitinv <- make.link('logit')$linkinv
panel2 <- function(x, y, ...){
  panel.superpose(x, y, 
                  type='l', ...)
  panel.xyplot(x, y, type='l', lwd=2, cex=0.6, ...)
}
parset <- simpleTheme(col=rep(1:4,rep(2,4)), lty=rep(1:2,4), lwd=2)
## c('solid','1141')
dat <- expand.grid(trtGp=factor(levels(HawCon$trtGp), levels=levels(HawCon$trtGp)),
                   TrtTime=pretty(range(HawCon$TrtTime),15), link=c('cloglog','logit'))
dat$scTime <- scale(dat$TrtTime)
dat$trtGpRep <- rep('new',nrow(dat))
hatClog <- predict(HCbb.cll, newdata=subset(dat, link=='cloglog'), se=TRUE, allow.new.levels=TRUE)
hatClog2 <- predict(HCbb2s.cll, newdata=subset(dat, link=='cloglog'), se=TRUE,  allow.new.levels=TRUE)
diffClog <- cloginv(hatClog2$fit)-cloginv(hatClog$fit)
hatLgt <- predict(HCbb.logit, newdata=subset(dat, link=='logit'), se=TRUE, allow.new.levels=TRUE)
hatLgt2 <- predict(HCbb2s.logit, newdata=subset(dat, link=='logit'), se=TRUE,  allow.new.levels=TRUE)
diffLgt <- logitinv(hatLgt2$fit)-logitinv(hatLgt$fit)
dat <- within(dat, {diff<-c(diffClog,diffLgt)
})
## dat <- within(dat, {lwr<-fit-2*se.fit; upr<-fit+2*se.fit})

gph <- xyplot(diff~TrtTime|link, outer=TRUE, data=dat, groups=trtGp,
              # upper = dat$upr, lower = dat$lwr, 
              panel = panel2,  
              xlab="Days in coolstorage", ylab="Difference in fitted value", 
              auto.key=list(text=levels(HawCon$trtGp), columns=4, points=FALSE, lines=TRUE), 
              par.settings=parset, layout=c(2,1), 
              scales=list(x=list(at=c(2,6,10,14)), 
              y=list(relation='free'), alternating=c(1,1)),
              strip=strip.custom(factor.levels=
            c("A: Complementary log-log link", "B: Logit link")))
gph
```

## Diagnostic checks 

Randomized quantile residuals, with the plots that are available
in the `DHARMa` provide helpful diagnostic check. For any
residual, the corresponding quantile residual is the proportion
of residuals expected, under model assumptions, to be less than
or equal to it.  If the distributional assumptions are satisfied,
the quantile residuals should have a distribution that differs
only by statistical error from a uniform distribution.

The function `DHARMa::simulateResiduals()` provides a
convenient means to simulate enough sets of residuals (by
default, 250) to give a good sense, for each individual
observation, of the distribution.  These then provide a
reference distribution for calculation of _quantile_
residuals.  Residuals may be calculated allowing only for
the fixed effects (the default), or conditional on one or
more levels of random effects. If the model is correct,
residuals should be uniformly distributed irrespective of
the conditioning.  See `?DHARMa::simulateResiduals` for
details.

* For the data as a whole, the distribution of residuals
  can be checked by plotting the quantile residuals against
  the corresponding quantiles.
    + Departures from assumptions show a pattern of difference
    from the line `y=x` that is different from that for 
    normal distribution quantile-quantile plots.
*  A second check plots quantile residuals against quantiles of
    predicted values. Quantile regression is then used to fit curves
    at 25%, 50%, and 75% quantiles of the quantile residuals. If the
    model is correctly specified, these should all be, to within
    statistical error, horizontal lines.
* Plots against other explanatory variables provide added
  checks.

Do such deviations from assumptions as are present matter?  
A useful device is to simulate new 'observations' from the model, 
and check whether there is a difference of substance in the 
fitted values and parameter estimates.

```{r echo=FALSE}
pcheck <- suppressMessages(requireNamespace('DHARMa', quietly = TRUE))
yesDHARMa <- pcheck
if (!yesDHARMa) {
   message("The suggested package `DHARMa` is not available/installed.")
message("Code that requires this package will not be executed.")
}
```

Figure \@ref(fig:DHARMa) shows the diagnostic plots for the linear model
with a complementary log-log link. 

```{r cap3, echo=FALSE} 
cap3 <- "Panel A shows the quantile-quantile plot,
for the linear model with a complementary log-log link.
Panel B plots estimated quantiles against mortality, 
while Panel C plots estimated quantiles against total 
number, on a logarithmic scale."
```

```{r DHARMa, fig.width=3.5, fig.asp=0.95, bot=-1, top=1.5, out.width="32%", warning=0, fig.align='center', fig.show="hold", message=FALSE, echo=FALSE, fig.cap=cap3, eval=yesDHARMa}
oldpar <- par(mar=c(3.1,3.1,2.6,1.1), mgp=c(2.1, 0.5,0))
set.seed(29)
simRef <- DHARMa::simulateResiduals(HCbb.cll, n=250, seed=FALSE)
DHARMa::plotQQunif(simRef)
DHARMa::plotResiduals(simRef)
DHARMa::plotResiduals(simRef, form=log(HawCon$Total), xlab="log(Total)")
par(oldpar)
```

The quantile-quantile (Q-Q) plot looks fine, The quantile residuals
from the data appear, if anything, closer to uniformly distributed than
any of the simulated sets of residuals. In Panels B and C, the quartiles
of the data are appear satisfactorily close to the relevant quartiles.


```{r cap4, echo=FALSE}
cap4 <- "Diagnostic plots, for the model with a logit link."
```

```{r residBYgp, fig.width=3.75, fig.asp=0.95, bot=-1, top=1.5, out.width="32%", warning=0, fig.align='center', fig.show="hold", message=FALSE, echo=FALSE, mar=c(3.1,3.1,2.6,1.1), fig.cap=cap4,eval=yesDHARMa}
oldpar <- par(mar=c(3.1,3.1,2.6,1.1), mgp=c(2.1, 0.5,0))
simResBB <- suppressWarnings(DHARMa::simulateResiduals(HCbb.cll, n=250) )
DHARMa::plotQQunif(simResBB)
DHARMa::plotResiduals(simResBB, xlab= "Predictions, Complementary log-log model")
simResLGT <- suppressWarnings(DHARMa::simulateResiduals(HCbb.logit, n=250) )
DHARMa::plotResiduals(simResLGT, xlab= "Predictions, logit model")
par(oldpar)

## Alternative -- group names are not shown:
## plotResiduals(simRes, form=HawCon$trtGp)
```

Now compare (Figure \@ref(fig:residBYgp)) scaled residuals between 
treatment groups.

```{r cap4.5, echo=FALSE}
cap4.5 <- "Quantile residuals, by treatment group, for the 
betabinomial model"
```

```{r trtGp, echo=FALSE, fig.width=5, fig.asp=0.7, bot=-2.5, warning=FALSE, fig.align='center', fig.show="hold", message=FALSE, warning=FALSE, out.width="49%", echo=FALSE, fig.cap=cap4.5, eval=yesDHARMa}
dotplot(scaledResiduals~HawCon$trtGp, data=simResBB, 
        scales=list(x=list(rot=30)), ylab="Quantile Residuals",
       main=list(expression(plain("A: Residuals, by treatment group")),
                      x=0.05, y=-0.2, just="left"))
bwplot(scaledResiduals~HawCon$trtGp, data=simResBB,  ylab="",
       scales=list(x=list(rot=30)),
       main=list(expression(plain("B: Boxplot comparison of residuals")),
                 x=0.05, y=-0.2, just="left"))
```



The numbers for `MelonEgg`, `MedL1`, and `MelonL1` are too small
to give useful boxplot displays.  Except for `MedEgg`, where
points are concentrated in the tails, the scatter of points in 
Panel A appears reasonably comparable between treatment groups.

QQ plots look good for all the models. In the sequel, they will
be left out.

### Uniform quantile-quantile plots --- an example. {-}

We will generate data from a highly overdispersed binomial type
distribution, then examine the uniform quantile-quantile plot
given by `DHARMa::plotResiduals()`.  An easy way to generate
overdispersed binomial type data is to start with binomial data,
then multiply both "successes" and "failures" by a number that
is substantially greater than one.

```{r cap4.75}
cap4.75 <- paste0("Diagnostics for model fitted to strongly overdispersed
binomial type data. Notice that the overdispersion results in an
S-shaped distribution of the residuals around the line $y=x$. The
boxplot is, in this case, uninformative.")
```

```{r overdispSim, echo=FALSE, fig.width=5, fig.asp=0.85, bot=-2.5, warning=FALSE, fig.align='center', fig.show="hold", out.width="49%", echo=FALSE, fig.cap=cap4.75, eval=yesDHARMa}
yes <- rbinom(n=100, size=50, prob=0.5)
sim <- cbind(yes=yes*4, no=200-yes*4)
sim.TMB <- glmmTMB::glmmTMB(sim~1, family=binomial)
sim250 <- DHARMa::simulateResiduals(sim.TMB)
DHARMa::plotQQunif(sim250)
DHARMa::plotResiduals(sim250)
```

### AIC-based model comparisons {-}

Comparisons should be for models with the same outcome
variable, and with the same 
data.^[See <https://robjhyndman.com/hyndsight/aic/> for a detailed commentary>]

In addition to the models described so far, we will
include also models, to be discussed below, that
assume binomial errors, with observational level
random effects added.
The model with the lowest AIC is the preferred model.

```{r obslevel, echo=FALSE, eval=TRUE}
ctl <- glmmTMB::glmmTMBControl(optimizer=optim,
                               optArgs=list(method="BFGS"))
HCbiObs.cll <- 
  glmmTMB::glmmTMB(cbind(Dead, Live) ~ 0 + trtGp/scTime +
                       (1|obs) + (scTime|trtGpRep),
                       family=binomial(link='cloglog'),
                       control=ctl, data=HawCon)
HCbiObs.logit <- 
  glmmTMB::glmmTMB(cbind(Dead, Live) ~ 0 + trtGp/scTime +
                          (1|obs) + (scTime|trtGpRep),
                      family=binomial(link='logit'),
                      control=ctl, data=HawCon)
```

The following shows AIC values
```{r glmmTMB-altFits-AIC, echo=FALSE}
aicStats <- 
  AIC(HCbb.cll,HCbb2s.cll,HCbb.logit, HCbb2s.logit, 
              HCbiObs.cll, HCbiObs.logit)
rownames(aicStats) <- substring(rownames(aicStats),3)
round(t(aicStats),2)
```

There is a strong preference for a complementary log-log
link, with the linear fit slightly better than the degree
2 normal spline fit, arguing for the use of the fitted
lines for calculation of LT99s and confidence intervals.

## Estimates of $\rho$, and of the dispersion factor

```{r cap8, echo=FALSE}
cap8 <- "Panels A and B show intra-class correlation estimates 
for, respectively, a complementary log-log link and a logit link. 
Both models assume a betabinomial error. Panel C shows, for 
the complementary log-log model, the dispersion factors that 
result."
```

```{r glmmTMB-altFits-gph-disp, fig.width=9, fig.height=3.5, top=2, out.width="100%",  fig.align='center', fig.show="hold", fig.cap=cap8, echo=FALSE}
oldpar <- par(oma=c(0,0,2,0))
parset <- simpleTheme(col=rep(1:4,rep(2,4)),pch=rep(c(1,2), 4), lwd=2)
HawCon$rho2clog <- qra::getRho(HCbb.cll)
HawCon$dispClog <- with(HawCon, 1+(Total-1)*rho2clog)
titles=c(expression("A: "*rho*", cloglog link"),expression("B: "*rho*", logit link"),
         "C: Dispersion, cloglog link")
library(lattice)
HawCon$rho2logit <- qra::getRho(HCbb.logit)
xyplot(rho2clog+rho2logit+dispClog ~ TrtTime, groups=trtGp, data=HawCon,
       outer=TRUE, between=list(x=0.25),
       par.settings=parset,
       scales=list(x=list(alternating=FALSE), y=list(relation='free',tick.number=4)),
       auto.key=list(columns=4, between.columns=2, between=1),
       xlab="Days in coolstorage", ylab="Parameter Estimate",
       strip=strip.custom(factor.levels=titles))
par(oldpar)
```

Figures \@ref(fig:glmmTMB-altFits-gph-disp)A and B (for a logit link)
plot the estimates of \(\rho\), based on modeling the
logarithm of the scale parameter as a degree 2 natural spline
function of `scTime` that is added to a straight line response 
that is different for each different treatment group. Use of
a logarithmic link (the default) has the consequence that 
effects that are additive on the link scale become multipliers 
when unlogged. 

Figure \@ref(fig:glmmTMB-altFits-gph-disp)C shows, for the 
complementary log-log model, the dispersion factors that 
result --- high at midrange times and mortalities, reducing
to close to a constant 1.0 at either extreme.  Use of a 
normal spline basis helps ensures that the value for $\rho$,
and hence the dispersion factor, extrapolates to a close
to constant value at both extremes. 

A dispersion factor that is close to 1.0 at the upper
extreme of the data provides a limited justification 
for assuming a binomial distribution, for purposes of 
calculating the sample size needed to demonstrate, at 
a 95% confidence level, a mandated survival rate of, 
e.g., no more than one in perhaps 100,000 insects.

# Binomial errors, plus observation level random effects

The suggestion here is that for the replicates within each
species/lifestage combination, the response in any one 
observation is close to binomial. Added to this is random 
variation between observations.  By comparison with fitting
a betabinomial or a quasibinomial error, the effect is to
reduce the variance of the observed mortalities at low and high 
mortality points, relative to variance at midrange values.

The following fits the two models:
```{r obslevel, echo=FALSE, eval=TRUE}
```

```{r cap2}
cap2 <- "AICs are compared between models with binomial family
errors and observation level random effects.  The two sets of
points make the comparison for, respectively, data that have 
been simulated from a model with logit link and a model with 
complementary log-log link.  The large triangle makes the
comparison for the models fitted to the 'HawCon' data."
```

```{r only-obslevel, fig.width=3.75, fig.asp=0.85, bot=-1, top=1.5, out.width="55%",  fig.align='center', fig.show="hold", message=FALSE, echo=FALSE, mar=c(3.1,3.1,2.6,1.1), fig.cap = cap2}
## Notice the use of the 'BFGS' optimization method in place 
## of the default.  
aicData <- setNames(AIC(HCbiObs.logit,HCbiObs.cll)[,2], c('logit','cll'))
set.seed(17)
sim <- simulate(HCbiObs.logit, nsim=10)
simcll <- simulate(HCbiObs.cll, nsim=10)
aic.cll <- aic2.cll <- aic.logit <- aic2.logit <- numeric(10)
for (i in 1:10){
  zlogit <- 
    glmmTMB::glmmTMB(sim[[i]] ~ 0 + trtGp/scTime +
                 (1|obs) + (1|trtGpRep),
             family=binomial(link='logit'),
             data=HawCon) 
  aic.logit[i] <- AIC(zlogit)
  zcll <- 
    glmmTMB::glmmTMB(sim[[i]] ~ 0 + trtGp/scTime +
                 (1|obs) + (1|trtGpRep),
             family=binomial(link='cloglog'),
             data=HawCon) 
  aic.cll[i] <- AIC(zcll)
  zlogit2 <- 
    glmmTMB::glmmTMB(simcll[[i]] ~ 0 + trtGp/scTime +
                 (1|obs) + (1|trtGpRep),
             family=binomial(link='logit'),
             data=HawCon) 
  aic2.logit[i] <- AIC(zlogit2)
  zcll2 <- 
    glmmTMB::glmmTMB(simcll[[i]] ~ 0 + trtGp/scTime +
                 (1|obs) + (1|trtGpRep),
             family=binomial(link='cloglog'),
             data=HawCon) 
  aic2.cll[i] <- AIC(zcll2)  
}
gph1 <- xyplot(aic.cll~aic.logit,
               key=list(text=list(c("logit model","cll model", "Data")), 
               points=list(pch=c(1,1,2),
                           col=c('blue','red','black')),columns=3))
gph2 <- xyplot(aic2.cll~aic2.logit, col='red')
gph3 <- gph1+latticeExtra::as.layer(gph2)+
  latticeExtra::layer(panel.abline(0,1), 
          panel.points(aicData[['logit']],
                       aicData[['cll']], pch=2, cex=2, col=1))
update(gph3, xlim=range(c(aic.logit, aic2.logit,
                          aicData['logit']))*c(.98,1.02), 
                        ylim=range(c(aic.cll, aic2.cll, 
                                     aicData['cll']))*c(.98,1.02))
```
Notice that the random effects term `(scTime|trtGpRep)` has been changed
to `(1|trtGpRep)`. This avoids convergence failure messages.

The AIC statistic shows a slight preference for the complementary
log-log model.  The comparison with data that have been simulated
from the respective fitted distributions indicates that the distinction
is far from clear.

The attempt to repeat a similar comparison with models that assume a
betabinomial error failed.  Most simulationa generated error messages.

Figure \@ref(fig:biObsCLL) compares the diagnostic plots, between
use of a complementary log-log link, and use of a logit link.

```{r cap5, echo=FALSE}
cap5 <- "Diagnostics for model with binomial errors and observation level
random effects."
```

```{r biObsCLL, fig.width=3.75, fig.asp=0.95, bot=-1, top=2.5, out.width="49%",  fig.align='center', fig.show="hold", message=FALSE, warning=0, echo=FALSE, mar=c(3.1,3.1,2.6,1.1), fig.cap=cap5, eval=yesDHARMa}
set.seed(29)
simRefcll <- suppressWarnings(
  DHARMa::simulateResiduals(HCbiObs.cll, n=250, seed=FALSE) )
DHARMa::plotResiduals(simRefcll, xlab='cll: model prediction')
DHARMa::plotResiduals(simRefcll, form=log(HawCon$Total), 
              xlab="cll: log(Total)")
simReflogit <- suppressWarnings(
  DHARMa::simulateResiduals(HCbiObs.logit, n=250, seed=FALSE) )
DHARMa::plotResiduals(simReflogit, xlab='logit: model prediction')
DHARMa::plotResiduals(simReflogit, form=log(HawCon$Total), 
              xlab="logit: log(Total)")
```

There appears to be substantial heteroscedasticity that is
not accounted for.  If the observation specific random error
could be modeled as a function of treatment time, it appears
likely that this model would do just as well, or possibly
even better, than the betabinomial model, with a model also
for the dispersion factor.

Code is

```{r biObsCLL, eval=FALSE, echo=TRUE}
```

# Confidence intervals for ratios

Fieller's formula provides a methodology for calculating
confidence intervals for ratios.  Here, it will be turned
to use for calculation of confidence intervals for exposures
(times, or doses) required to kill 99% of insects.

## 99\% Lethal time estimates and confidence intervals

The estimated time required to kill 99% of insects
(lethal time 99%, or LT99) is commonly used as a starting point
for assessing what time might be effective in commercial
practice.  Thus, for the model that used a complementary 
log-log link, and setting:
$$
y = \log(-\log(1-0.99)) = 1.53,
$$
one solves for
$x = \mbox{LT99}$ in an equation of the form $y = a + b x$.
Thus:
$$
\mbox{LT99} = x = \dfrac{y - a}{b}
$$

The determination of confidence intervals for such ratios
is one of a much wider class of problems.  The Fieller's
formula approach (@morgan_1992), implemented in the
`qra` package, makes the common assumption that 
$(y-a,\: b)$ has been sampled from a distribution for which
the bivariate normal is an adequate approximation.
See `?qra::fieller`. 

The sampling distribution of the calculated value $x$
is typically, unless $\mbox{var}[b]$ is small relative to $b$,
long-tailed.  As a consequence, the *Delta* method, which
uses an approximate variance as the basis for inference, is in
general unreliable.  

The Fieller's formula approach cannot in general be adapted 
to give confidence intervals for differences of LT99s or 
other such ratio statistics, unless the denominators of 
the statistics that are compared happen to be the same. 
A usable implementation of the simulation approach, 
which seems needed for the calculation of confidence
intervals for LT99 differences, can in principle be handled
using the function `lme4::bootMer()`.

## What difference does the choice of model make?

We now proceed to investigate the damage done by assuming
a constant dispersion factor, or by specifying a binomial
error (there is no allowance for a dispersion factor), or
by use of a logit link in place of a betabinomial link.
Figure \@ref(fig:plotCI) compares confidence intervals, calculated
using Fieller's formula, from models whose confidence intervals
are identified thus:

* `BB-cll`, noting that `cll=cloglog`, identifies the preferred
model, saved as `HCbb.cll`.  The dispersion parameter is modeled 
as a degree 2 normal spline function of treatment time.\label{item:1}
* `BIobsRE-cll`, saved as `HCbiObs.cll`, which adds observation
level random effects to a binomial error.
* `BB-logit`, saved as `HCbb.logit`, 
replaces the complementary log-log link of item 1
with a logit link, while retaining the same 
form of degree 2 normal spline in `trtTime` for 
the dispersion parameter.
    + All estimates are then shifted upwards, in four cases with
    a much higher upper limit than for the preferred model.
* `BB-cll, const Disp factor` identifies a model that differs
from `HCbb.cll` by fitting a constant dispersion factor; 
    + This leads, in 7 out of 8 treatment groups, to a wider 
    confidence interval.
* `Binomial-cll`, saved as `HCbin.LTcll` assumes 
binomial errors, with overdispersion accounted for in the between 
treatment component of variance;
    + The transfer of all extra-binomial variance to the between
    treatments component leads, for the treatment group `MelonL1`,
    to a dramatic increase in the confidence interval width.

In the two models (`HCbb.cll` and `HCbb.logit`) that assume
a betabinomial error and model the variation in the variance, 
the dispersion factor parameter determines the
intra-class correlation \(\rho\). 

```{r shorten, echo=FALSE}
shorten <- function(nam, leaveout=c('trtGp','Fly',':')){
  for(txt in leaveout){
    nam <- gsub(txt,'', nam, fixed=TRUE)
  }
  nam
}
```

```{r crude-cll, echo=FALSE, warning=F}
## Fit two simplistic and unsatisfactory models.
HCbbDisp1.cll <- update(HCbb.cll, dispformula=~1)
HCbin.cll <- update(HCbb.cll, family=binomial(link="cloglog"))
```


```{r extract-BB-LTcll, echo=FALSE}
LTbb.cll <- qra::extractLT(p=0.99, obj=HCbb.cll, link="cloglog",
                              a=1:8, b=9:16, eps=0, df.t=NULL)[,-2]
rownames(LTbb.cll) <- shorten(rownames(LTbb.cll))
```

```{r extract-BI-obsRE, echo=FALSE}
## NB: The formula has used the scaled value of time.
## Below, `offset` is used to retrieve the scaling parameters 
## `center` ## and `scale` in `(x-center)/scale`.
offset <- qra::getScaleCoef(HawCon$scTime)
LTbiObs.cll <- qra::extractLT(p=0.99, obj=HCbiObs.cll,
                          a=1:8, b=9:16, eps=0, offset=offset,
                          df.t=NULL)[,-2]
rownames(LTbiObs.cll) <- shorten(rownames(LTbiObs.cll))
```


```{r extract-BB-LTlogit, echo=FALSE}
LTbb.logit <- qra::extractLT(p=0.99, obj=HCbb.logit, link="logit",
                          a=1:8, b=9:16, eps=0, offset=0,
                          df.t=NULL)[,-2]
rownames(LTbb.logit) <- shorten(rownames(LTbb.logit))
```


```{r extract-crude-LTcll, echo=FALSE}
LTbbDisp1.cll <- 
  qra::extractLT(p=0.99, obj=HCbbDisp1.cll, 
                 a=1:8, b=9:16, eps=0, df.t=NULL)[,-2]
rownames(LTbbDisp1.cll) <- shorten(rownames(LTbbDisp1.cll))
LTbin.cll <- 
  qra::extractLT(p=0.99, obj=HCbin.cll, 
                 a=1:8, b=9:16, eps=0, df.t=NULL)[,-2]
rownames(LTbin.cll) <- shorten(rownames(LTbin.cll))
```

```{r OKplotrix-plotCI}
OKplotrix <- suppressWarnings(requireNamespace("plotrix", quietly = TRUE))
if (!OKplotrix) {
   message("The `plotix' package is not available/installed.")
message("Code that requires the `plotix' package will not be executed.")
}
```

```{r cap9, echo=FALSE}
cap9 <- "LT99 $95\\%$ confidence intervals are compared between
five different models."
```

```{r plotCI, echo=FALSE, fig.width=7.0, bot=1.0, fig.asp=0.625, warning=FALSE, fig.align='center', message=FALSE, out.width="75%", echo=FALSE, fig.cap=cap9, eval=OKplotrix}
gpNam <- rownames(LTbb.cll)
ordEst <- order(LTbb.cll[,1])
col5 <- c("blue","lightslateblue","brown",'gray','gray50')
plotrix::plotCI(1:8-0.34, y=LTbb.cll[ordEst,1], ui=LTbb.cll[ordEst,3],
       li=LTbb.cll[ordEst,2], lwd=2, col=col5[1], xaxt="n", 
       fg="gray", xlab="", ylab="LT99 Estimate (days)", 
       xlim=c(0.8,8.2), ylim=c(0,29), sfrac=0.008)
plotrix::plotCI(1:8-0.17, y=LTbiObs.cll[ordEst,1], ui=LTbiObs.cll[ordEst,3],
       li=LTbiObs.cll[ordEst,2], lwd=2, col=col5[2], xaxt="n", 
       fg="gray", xlab="", ylab="LT99 Estimate (days)", 
       xlim=c(0.8,8.2), ylim=c(0,29), add=TRUE, sfrac=0.008)
plotrix::plotCI(1:8, y=LTbb.logit[ordEst,1], ui=LTbb.logit[ordEst,3],
       li=LTbb.logit[ordEst,2], lwd=2, col=col5[3], xaxt="n", 
       add=TRUE, sfrac=0.008)
plotrix::plotCI(1:8+0.17, y=LTbbDisp1.cll[ordEst,1], ui=LTbbDisp1.cll[ordEst,3],
       li=LTbbDisp1.cll[ordEst,2], lwd=2, col=col5[4], xaxt="n", 
       add=TRUE, sfrac=0.008)
plotrix::plotCI(1:8+0.34, y=LTbin.cll[ordEst,1], ui=LTbin.cll[ordEst,3],
       li=LTbin.cll[ordEst,2], lwd=2, col=col5[5], xaxt="n", 
       add=TRUE, sfrac=0.008)
axis(1, at=1:8, labels=gpNam[ordEst], las=2, lwd=0, 
     lwd.ticks=0.5, col.ticks="gray")
legend("topleft", legend=c("BB-cll (cll=cloglog)",  "BB-cll-ObsRE", "BB-logit",
                           "BB-cll, const Disp factor", 
                           "Binomial-cll"),
       inset=c(0.01,0.01), lty=c(1,1), col=col5[1:5],
       text.col=col5[1:5], bty="n",y.intersp=0.85)
```

Code to extract the 95% confidence interval for the LT99 is:
```{r extract-BB-LTcll, eval=FALSE, echo=TRUE}
```

Other confidence intervals are calculated by modifying the
call to `extractLT()` as required.

The message from Figure \@ref(fig:plotCI) is that, for
purposes of estimating the LT99 or other high lethal time
point, assumptions for the error family as well as for the 
link can make a large difference. Where those choices are
made casually, without careful checking, serious biases 
may result.  Use of a model that is unsatisfactory 
within the range of the available data is not a good
starting point for extrapolation into high mortality 
regions, with the additional uncertainties that then 
result.

# Further models and model fitting functions

## Binomial errors, observation level random effects, and more

This adds to the earlier observation level random effects model
by allowing the associated variance to differ between
species/lifestage/replicate combinations.  Contributions to 
the variance at all levels other than the observation add to 
the binomial variance.  

The code that follows implements such a model.  It does not
improve on the observation level random effects model demonstrated
earlier.

```{r obsLevel1, echo=FALSE, eval=FALSE}
dMedEgg <- with(HawCon, dummy(trtGp,"MedFlyEgg:"))
dMedL1 <- with(HawCon, dummy(trtGp,"MedFlyL1:"))
dMedL2 <- with(HawCon, dummy(trtGp,"MedFlyL2:"))
dMedL3 <- with(HawCon, dummy(trtGp,"MedFlyL3:"))
dMelEgg <- with(HawCon, dummy(trtGp,"MelonFlyEgg:"))
dMelL1 <- with(HawCon, dummy(trtGp,"MelonFlyL1:"))
dMelL2 <- with(HawCon, dummy(trtGp,"MelonFlyL2:"))
dMelL3 <- with(HawCon, dummy(trtGp,"MelonFlyL3:"))
formXre <- cbind(Dead, Live) ~ 0 + trtGp/scTime +
  (1|obs) + (1|trtGpRep) + 
  (0 + dMedEgg| trtGpRep) + (0 + dMedL1 | trtGpRep) +
  (0 + dMedL2 | trtGpRep) + (0 + dMedL3 | trtGpRep) +
  (0 + dMelEgg| trtGpRep) + (0 + dMelL1 | trtGpRep) +
  (0 + dMelL2 | trtGpRep) + (0 + dMelL3 | trtGpRep)
## NB: The formula has used the scaled value of time.
## Below, `offset` is used to record the scaling parameters 
## `center` ## and `scale` in `(x-center)/scale`.
offset <- with(attributes(HawCon$scTime),
               c(`scaled:center`, `scaled:scale`))
HCXre.biObs <- gkmmTMB::glmmTMB(formXre, family=binomial(link='cloglog'),
                       control=ctl, data=HawCon)
## Could not get the following to converge
## formXreS <- update(formXre, ~ .+ trtGp/splines::ns(scTime,2))
```

## Fits using `lme4::glmer()`

This is an alternative to `glmmTMB()`, for fits that
assume a binomial error, plus observation level random
effects.  The AIC statistics differ slightly from 
those given using `glmmTMB()`, for reasons that are
unclear.

```{r glmerFits}
## Comparisons using `glmer()`
HCglmerBIobs.cll <- 
  lme4::glmer(cbind(Dead, Live) ~ 0 + trtGp/scTime + (1|obs) + (1|trtGpRep),
        family=binomial(link='cloglog'), nAGQ=1, data=HawCon,
        control=lme4::glmerControl(optimizer='bobyqa'))
HCglmerBIobs2s.cll <- suppressWarnings(
  lme4::glmer(cbind(Dead, Live) ~ 0 + trtGp/scTime + splines::ns(scTime,2) +
        (1|obs) + (1|trtGpRep), family=binomial(link='cloglog'),
        nAGQ=0, data=HawCon))
HCglmerBIobs.logit <- lme4::glmer(cbind(Dead, Live) ~ 0 + trtGp/scTime +
                           (1|obs) + (1|trtGpRep),
        family=binomial(link='logit'), nAGQ=0, data=HawCon, 
        control=glmerControl(optimizer='bobyqa'))
HCglmerBIobs2s.logit <- 
  lme4::glmer(cbind(Dead, Live) ~ 0 + trtGp/scTime + splines::ns(scTime,2) +
        (1|obs) + (1|trtGpRep), family=binomial(link='logit'), nAGQ=0,
                       data=HawCon, control=glmerControl(optimizer='bobyqa'))
```

```{r cfAICs}
cfAIC <- 
  AIC(HCbiObs.cll,HCglmerBIobs.cll,HCbiObs.logit,HCglmerBIobs.logit,
             HCglmerBIobs2s.cll,HCglmerBIobs2s.logit)
rownames(cfAIC) <- c('TMB:cll','mer:cll','TMB:lgt','mer:lgt',
                     'mer:cllCurve', 'mer:lgtCurve')
(tab <- t(round(cfAIC,2)))
```

### AIC statistics {-}

The first two AIC values, for the models with a complementary 
log-log link, are `r tab[2,1]`
(for the `glmmTMB` model), and `r tab[2,2]` (for the `glmer` model).
The third and fourth values, both for the logit link, are `r tab[2,3]` 
(`glmmTMB`) and `r tab[2,4]` (`logit`).  In both cases, `glmer` models 
that allow for overall curvature give an increased AIC statistic.
The reason for the small difference between `glmer()` fits and `glmmTMB()`
fits is unclear.

### Control settings {-}

Use of `glmer()` with `nAGQ=1` results in an AIC statistic that is
closer to that from the `glmmTMB()` fit.  The `glmer()` default 
has `optimizer = c("bobyqa", "Nelder_Mead")`, which has the effect
of using the first optimizer for the first part of the calculation,
then moving to use the second optimizer.  See `?glmerControl`.
Setting `control=glmerControl(optimizer='bobyqa')`, i.e., stay
with "bobya" right through, avoids a "failed to converge" warning
that for this fit occur with the default.

This choice was made after using the function `allFit()` to 
run the fit with a range of range of methods and optimizers, then 
checking for possible differences in the loglikelihoods and in the 
parameter estimates.  Code used is, with the somewhat confusing
output omitted.

```{r allFit, results='hide', echo=TRUE, eval=FALSE}
check <- (requireNamespace('dfoptim',quietly=TRUE)&
    requireNamespace('optimx',quietly=TRUE))
if(check)
ss<-suppressWarnings(summary(allFit(HCglmerBIobs.cll)))
```

```{r names-ss, hold=FALSE, echo=-2, eval=FALSE}
stopifnot(check)
options(width=70)
names(ss)
ss$msgs
ss$llik
```

## Gaussian errors, on a complementary log-log scale

```{r LT99gauss.LTcll}
cloglog <- make.link('cloglog')$linkfun
HCgauss.cll <- glmmTMB::glmmTMB(cloglog((PropDead+0.002)/1.004)~0+
                         trtGp/TrtTime+(TrtTime|trtGpRep), 
                       family=gaussian(), data=HawCon)
LTgauss.cll <- qra::extractLT(p=0.99, obj=HCgauss.cll, link="cloglog",
                                 a=1:8, b=9:16, eps=0.002, offset=c(0,1),                                 df.t=NULL)[,-2]
rownames(LTgauss.cll) <- shorten(rownames(LTgauss.cll))
```

A further model that might be tried is a linear mixed model, 
with \(log(1-log((p+0.002)/(1+0.004)))\) as the dependent 
variable (complementary log-log link), and gaussian error.
Figure \@ref(fig:Gauss-simRes)  shows the plot of residuals versus
predicted quantile deviations.

```{r cap11, echo=FALSE}
cap11 <- "Residuals versus predicted quantile deviations, for
the linear mixed model, 
with \\(log(1-log((p+0.002)/(1+0.004)))\\) as the dependent 
variable, complementary log-log link, and gaussian error."
```

```{r Gauss-simRes, echo=FALSE, w=3.5, fig.asp=0.75, left=-0.5, bot=-1, mgp=c(3,1,0), crop=TRUE, fig.align='center', out.width="50%", fig.cap=cap11, eval=yesDHARMa}
sim <- DHARMa::simulateResiduals(HCgauss.cll)
DHARMa::plotResiduals(sim)
```

```{r cap12, echo=FALSE}
cap12 <- "Comparison of estimates and upper and lower $95\\%$ 
confidence limits, between the preferred betabinomial 
complementary log-log model (BB) and the gaussian error model."
```

```{r BBvsGauss, out.width="100%",  message=FALSE, echo=FALSE}
cfLTs <- cbind(LTbb.cll, LTgauss.cll)
colnames(cfLTs) <- c(rep('BB',3),rep('gauss',3))
tab <- round(cfLTs[,c(c(1,4),c(1,4)+1,c(1,4)+2)],1)
pcheck <- suppressWarnings(requireNamespace("kableExtra", quietly = TRUE))
if(pcheck) pcheck & packageVersion("kableExtra") >= "1.2"
if(pcheck){
kableExtra::kbl(tab, font_size=9, caption=cap12) |>
  kableExtra::kable_paper("striped", full_width=FALSE) |>
  kableExtra::column_spec(6:7, bold=TRUE) |>
  kableExtra::row_spec(5:8, color='blue', bold=T) |>
  kableExtra::add_header_above(c(' '=1,'Estimate'=2,'Lower'=2, 'Upper'=2), align='c') 
} else print(tab)
```

Table \@ref(tab:BBvsGauss) compares the LT99 95\% confidence interval
estimates and bounds.  Note the big differences for the Egg and L1 
(larval stage 1) results.

The issue here is that all points where the data show 100%
mortality transform to the same $y$-ordinate in the plot of
\(log(1-log((p+0.002)/(1+0.004)))\) against `TrtTime` .

# Parting comments

## The assumed dose-response is a rough approximation!

Quite strong assumptions, which could be checked only to a
limited extent, have been made in order to get the results
given. The checks that were performed made it clear that
that the model that waqs finally chosen was not quite correct.
The model that was chosen as "best" assumed that

* The pattern of mortality response is, on a complementary
log-log scale, linear with time, consistently across the eight
species/lifestage combinations.  
    + We did check whether it was
linear as opposed to a degree 2 normal spline, but the 
difficulties involved in getting a model to fit the limited 
data did not allow a check for a degree 3, or more complex, 
response pattern.  
* Within replicate errors follow a betabinomial distribution, 
albeit with a dispersion parameter that varies with time in 
coolstorage.
    + The modeling of variation in the dispersion
parameter ensures that, whether or not the 
betabinomial assumption is strictly correct, some limited 
account is taken of the pattern of change of variance with
time.  It is much less crude than assuming a binomial within
replicate error, and using between replicate variation to
account for extra-binomial variation, with no adjustment
for variation with time in coolstorage.  
* Results can be extrapolated to give the times needed
to give, with some limited confidence, 99% or higher
mortalities.  
    + In order to gain any reasonable
level of confidence that the model continues to give
acceptably accurate predictions at the times required
for 99% mortality levels and beyond, huge numbers of
insects, spread across a vastly more replicates than
in a study such as generated this dataset, would be
required.  
* Tolerance to treatment is constant within a lifestage.
    + This is unlikely to be strictly true.

Additionally, in the absence of provision for calculation
of the Kenward-Roger or other well-validated degrees of 
freedom approximation for models fitted using `glmmTMB()`
or `glmer()`, the default that is used by `extractLT()` 
has to treated as a rough approximation.  See @pbkrtest for 
details of the Kenward-Roger approximation.

The results given should, accordingly, be used with
caution. Common practice is to use such results to identify 
a "most tolerant" lifestage, and to suggest a treatment 
protocol, which is then be checked out in a large-scale 
trial.  

One could have better confidence in models that had been
checked out against a wide range of relevant data. For a
given response probability $\pi$, does the multiplier for
the binomial variance $n \pi (1-\pi)$ increase with $n$,
as use of a betabinomial error assumes?

A useful first step would be creation of an open database 
into which researchers would be required, or at least 
strongly encouraged, to place their data. This would 
allow checking of model predictions for each specific 
treatment type, and for each class of pathogen, against 
the times found to give high mortality in large-scale 
trials.  While there is an extensive literature that 
presents results of analyses from relevant trial data, 
and a very limited literature that makes comparisons 
across a number of different datasets, few of the 
relevant datasets are available in the public domain.

<!-- ## Slope Versus Intercept Plot -->

<!-- ```{r ab-cap} -->
<!-- cap11 <- "Slope versus intercept plot.  The intercept is a difference -->
<!-- from the mean. Contour lines have been added for constant LT99=11 days -->
<!-- (labeled with Es), LT99=9 days (labeled with 9s), and LT99=7 days -->
<!-- (labeled with 7s)." -->
<!-- ``` -->

<!-- Attention is limited to the linear mixed model. 90% confidence ellipses -->
<!-- are added. -->

<!-- ```{r HawCon-print, fig.width=6.5, fig.height=5, out.width="75%", message=FALSE, warning=FALSE, fig.cap=cap11, eval=TRUE} -->
<!-- ## Extract variance-covariance matrix for estimates -->
<!-- ellipseFun <- function(which.coef=c(3,nFits+3), center, varmat, level=0.9, nFits, dfd){ -->
<!--   rad <- sqrt(2 * qf(level, 2, dfd)) -->
<!-- car::ellipse(center=center[which.coef], -->
<!--              shape=varmat[which.coef,which.coef], -->
<!--              radius=rad, center.pch=1, col="gray") -->
<!-- } -->
<!-- ests <- fixef(modX.rlmer) -->
<!-- names(ests) <- substring(names(ests),6) -->
<!-- vv <- as.matrix(vcov(modX.rlmer)) -->
<!-- ab <- matrix(ests, ncol=2) -->
<!-- nFits <- nrow(ab) -->
<!-- rnam <- names(ests[1:nFits]) -->
<!-- dimnames(ab) <- list(rnam, c("Intercept", "Slope")) -->
<!-- plot(ab, xlab="Intercept", ylab="Slope") -->
<!-- posn <- 2 + 2*(ab[,1]<median(ab[,1])) -->
<!-- text(ab, labels=rnam, pos=posn) -->
<!-- ellipseFun(which.coef=c(7,nFits+7), center=ests, varmat=vv, nFits=nFits, dfd=2*nFits) -->
<!-- ellipseFun(which.coef=c(3,nFits+3), center=ests, varmat=vv, nFits=nFits, dfd=2*nFits) -->
<!-- ellipseFun(which.coef=c(4,nFits+4), center=ests, varmat=vv, nFits=nFits, dfd=2*nFits) -->
<!-- ellipseFun(which.coef=c(5,nFits+5), center=ests, varmat=vv, nFits=nFits, dfd=2*nFits) -->
<!-- xy <- data.frame(a = pretty(ab[,1],20)) -->
<!-- linkfun <- make.link('logit')[['linkfun']] -->
<!-- xy[['b11']] <- (-xy[[1]]+linkfun(0.99))/(11-ascale) -->
<!-- xy[['b9']] <- (-xy[[1]]+linkfun(0.99))/(9-ascale) -->
<!-- xy[['b7']] <- (-xy[[1]]+linkfun(0.99))/(7-ascale) -->
<!-- lines(xy[,c("a","b11")], lty=2, type="b", pch="E", col=2) -->
<!-- lines(xy[,c("a","b9")], lty=2, type="b", pch="9", col=2) -->
<!-- lines(xy[,c("a","b7")], lty=2, type="b", pch="7", col=2) -->
<!-- ``` -->

<!-- ## Hotelling T^2^ -->
<!-- The following uses a Hotelling T^2^ test to compare the means -->
<!-- in the cases where the two ellipses (medFly L1 and MedFly L3) -->
<!-- are closest to touching. The following requires careful checking. -->
<!-- The $p$-value should be adjusted upwards for the number of -->
<!-- comparisons made. -->
<!-- ```{r T2,eval=TRUE} -->
<!-- ## Variance-covariance matrix for difference of intercepts, -->
<!-- ## and difference of slopes -->
<!-- dmat <- matrix(0, nrow=2, ncol=2*nFits) -->
<!-- dmat[1,c(5, nFits+5)] <- 1 -->
<!-- dmat[2,c(4, nFits+4)] <- -1 -->
<!-- v <- dmat%*%vv%*%t(dmat) -->
<!-- d <- ests[c(5,nFits+5)]-ests[c(4,nFits+4)] -->
<!-- ## T2 statistic for comparing the two bivariate means -->
<!-- T2 <- as.vector(d%*%solve(v)%*%d) -->
<!-- ## Equivalent approx F-statistic -->
<!-- F2 <- T2 * 2*nFits/(2*nFits-2) -->
<!-- print(1-pf(F2,2,2*nFits-2), digits=3) -->
<!-- ``` -->

<!-- On Hotelling T^2^, see, e.g., -->
<!-- https://en.wikipedia.org/wiki/Hotelling%27s_T-squared_distribution#Statistic -->

# References

